---
title: "Workshop 8"
subtitle: "F 9-10:50am"
author: "Ben Wollack"
date: "2026/21/05"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?  

**Specific leaf area** is a continuous variable measured in cm^2^/g.  

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).  

**Species** is a categorical variable (again, a *factor*) with 7 levels.  

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| _Encelia californica_        | ENCCAL       | Bush sunflower     |  
| _Eschsholzia californica_    | ESCCAL       | California poppy   | 
| _Penstemon centranthifolius_ | PENCEN       | Scarlet bugler     | 
| _Grindelia camporums_        | GRINCAM      | Gumweed            | 
| _Salvia leucophylla_         | SALLEU       | Purple sage        | 
| _Stipa pulchra_              | STIPUL       | Purple needlegrass | 
| _Lotus scoparius_            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here
drought_exp <- read_xlsx(
  here("data",
       "Valliere_etal_EcoApps_Data.xlsx"),
  # use this to specify sheet within excel
  sheet = "First Harvest"
)
```

```{r}
#| label: plot colors and aesthetics

# saves it for all future plots

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

plant_colors <- c(
  "Lotus scoparius" = lotsco_col,
  "Penstemon centranthifolius" = pencen_col,
  "Salvia leucophylla" = salleu_col,
  "Encelia californica" = enccal_col,
  "Stipa pulchra" = stipul_col,
  "Eschscholzia californica" = esccal_col,
  "Grindelia camporum" = gricam_col
)

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```


# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? [Dataset]. Dryad. [https://doi.org/10.5061/dryad.v0861f7](https://doi.org/10.5061/dryad.v0861f7)

```{r}
#| label: data cleaning

# create new object called drought_exp_clean
drought_exp_clean <- drought_exp |> 
  # lowercase letters in columns, replace spaces with underscores
  clean_names() |> 
  # create new column with species full names
  mutate(species_name = recode_values(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # create new column with descriptive water treatment names
  mutate(water_treatment = recode_values(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  )) |> 
  # set levels for linear model using species as a factor, 
  # relevel allows us to choose levels within factor, reference is lotus scoparius
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  # do same thing for water treatment
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # keep relevant columns: species, water treatment, specific leaf area, plant mass
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.  

```{r}
#| label: check-structure

str(drought_exp_clean)
```

Factors and levels display properly, so we're set for the linear model. `This is the only way to check.`

Insert code to display 10 random rows from the data frame here.  

```{r}
#| label: display-random

# display 10 rows from data frame
slice_sample(
  drought_exp_clean, 
  n = 10
  )
```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?  

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.  

Label the axes with units where appropriate.  

Add a title describing the relationship (e.g. as SLA increases, total mass ______). 

```{r}
#| label: scatterplot-sla

# base layer for scatterplot
# relating specific leaf area (x) to total mass (y)
ggplot(
  data = drought_exp_clean,
  mapping = aes(
    x = sla,
    y = total_g,
  )
) +
  # add points
  geom_point() +
  # rename axes, add title, the \U00B/ is unicode for a superscript
  labs(
    x = "Specific Leaf Area (cm\U00B2/g)",
    y = "Total Mass (g)",
    title = "As specific leaf area increases, total mass increases"
  )
```

## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: jitterplot-water

# create base layer
# relate water treatment to total mass
ggplot(
  data = drought_exp_clean,
  mapping = aes(
    x = water_treatment,
    y = total_g,
    color = water_treatment
  )
) + 
  # create jitterplot, remove vertical jitter and narrow horizontal jitter
  geom_jitter(height = 0,
              width = 0.2) +
  # change colors using colors set earlier
  scale_color_manual(
    values = c("Drought stressed" = ds_col,
               "Well watered" = ww_col)
  ) +
  # rename axes, create title
  labs(
    x = "Water Treatment",
    y = "Total Mass (g)",
    title = "Well watered plants have higher mass"
  ) +
  # remove legend
  theme(legend.position = "none")
  
```

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.  

Label the axes with units where appropriate.  

Add a title describing the differences in (mean or median) plant mass between species.  

Bonus: order the x-axis by decreasing mean plant mass (no, I like the order it's in and I don't want to mess up the reference species)

```{r}
#| label: jitter-species

# create base layer
# relate water treatment to total mass
ggplot(
  data = drought_exp_clean,
  mapping = aes(
    x = species_name,
    y = total_g,
    color = species_name
  )
) + 
  # create jitterplot, remove vertical jitter and narrow horizontal jitter
  geom_jitter(height = 0,
              width = 0.2) +
  # rename axes, create title
  labs(
    x = "Species",
    y = "Total Mass (g)",
    title = "Grindelia has the highest mass and Lotus has \nthe lowest"
  ) +
  # remove legend
  # hjust ensures the text does not bleed into the chart
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1)) +
  # use plant colors set before
  scale_color_manual(values = plant_colors) +
  # wrap text to prevent long text strings
  scale_x_discrete(labels = label_wrap(10))
```

# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species | Predictor list              |  
|--------------|:---:|:---------------:|:-------:|:---------------------------:|  
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |       X         |    X    | all predictors (full model) | 
| 2            |  X  |       X         |         | SLA and water treatment     |   
| 3            |  X  |                 |    X    | SLA and species             |    
| 4            |     |       X         |    X    | water treatment and species |    
| 5            |  X  |                 |         | SLA                         |    
| 6            |     |       X         |         | water treatment             |    
| 7            |     |                 |    X    | species                     |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model
model0 <- lm(
  total_g ~ 1, # formula no predictors
  data = drought_exp_clean # use cleaned data
)

# model 1: all predictors
model1 <- lm(
  total_g ~ sla + water_treatment + species_name, # formula all predictors
  data = drought_exp_clean # use cleaned data
)

# model 2: SLA and water treatment
model2 <- lm(
  total_g ~ sla + water_treatment, # formula sla + water
  data = drought_exp_clean # use cleaned data
)

# model 3: SLA and species
model3 <- lm(
  total_g ~ sla + species_name, # formula sla + species
  data = drought_exp_clean # use cleaned data
)

# model 4: water treatment and species
model4 <- lm(
  total_g ~ water_treatment + species_name, # formula water + species
  data = drought_exp_clean # use cleaned data
)

# model 5: SLA
model5 <- lm(
  total_g ~ sla, # formula sla
  data = drought_exp_clean # use cleaned data
)

# model 6: water treatment
model6 <- lm(
  total_g ~ water_treatment, # formula water
  data = drought_exp_clean # use cleaned data
)

# model 7: species
model7 <- lm(
  total_g ~ species_name, # formula species
  data = drought_exp_clean # use cleaned data
)
```

Dredge does this automatically, but avoid this because it causes you to avoid thinking about your data.

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.  

Arrange by decreasing AIC.

```{r}
#|label: calc-AIC

# put all models into calculating AICc
AICc(
  model0,
  model1,
  model2,
  model3,
  model4,
  model5,
  model6,
  model7
) |> 
  # sorts the AICc column by the values in that column
  arrange(AICc)
  # this shows that model 4 is the best model (best by >2)
  # this is water treatment and species
```

## Model Diagnostics

Check for best model

```{r}
#| label: check-diagnostics

# display in 2x2 grid
par(mfrow = c(2, 2))
# display plots
plot(model4)
```

Clumps come from the fact it's categorical, not continuous. There's no vertical patterns.

Residuals are normally distributed and homescedastic, and there's no outliers.

Insert code to display the summary for the best model. 

```{r}
#| label: model-summary

# display summary for model 4
summary(model4)
```

Intercept: model prediction for plant mass for Lotus scoparius in drought stressed treatment (these are the two reference levels)

The other treatments/species predict how much higher plant mass will be if that category is present.

Ex: The well watered treatment has an estimate of 0.0546 with a standard error of 0.0245. This is the difference in Lotus scoparius mass between the drought stressed and well watered treatment

Penstemon centranthifolius has one of 0.05 with a standard error of 0.0324, note that p > 0.05, so this may be 0.

## Stop and think: what does this model mean?  

**What is the best model?**

The best model includes water treatment and species as predictors

**How much variation in the response (total mass, in grams) does this model explain?**

Adjusted R-squared is 0.726, so it's 72.6% of the variation

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

Water treatment and species influence plant mass

# 4. Model predictions

Insert code to generate model predictions for the best model.  

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

```{r}
#| label: model-predictions

preds <- ggpredict(
  model4,
  terms = c("water_treatment", "species_name")
) |> 
  rename(water_treatment = x,
         species_name = group)
```


# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

```{r}
#| label: final-fig

ggplot(data = drought_exp_clean,
       mapping = aes(
         x = water_treatment,
         y = total_g,
         color = water_treatment
       )) +
  geom_jitter(
    height = 0,
    width = 0.1,
    shape = 21,
    alpha = 0.8
  ) +
  geom_pointrange(data = preds,
                  mapping = aes(
                    x = water_treatment,
                    y = predicted,
                    ymin = conf.low,
                    ymax = conf.high
                  )) +
  facet_wrap(~ species_name) +
  theme(legend.position = "none")
```



# 6. Writing

## Figure caption

## Basic components of test

**predictors**  


**response**  


**test**  


**distribution**  


**parameters (degrees of freedom)**  


**R^2^**  


**p-value**   


**significance level**  


**reference level(s)**  


## Written communication

### Methods

We used linear models to determine the predictive relationship between specific leaf area (cm^2^/g), treatment, and species on plant mass (g). We evaluated model assumptions of normally distributed errors using QQ plots of model residuals. We evaluated assumptions of constant variance of errors using Residuals vs Fitted and Scale-Location plots. We used Akaike's Information Criterion (AIC) to select the simplest model that explained the most variation in plant mass.  

### Results

We found that the simplest model describing the greatest variation in plant mass included treatment and species as predictors (linear model: F(2, 67) = 14.57, p < 0.001, $\alpha$ = 0.05). 







